Trail Saving on Backtrack

### 2020-02-16 figures/cactus_cadical.tex
fb notes for changes

\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt_data.csv};

\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-chrono-dl,maplelcmdist-trail-dl}
\end{axis}
\end{tikzpicture}

>>>>

\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-chrono_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-trail-working-version_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_cadical-master-trail-1lvl-proper_data.csv};

\legend{cadical,cadical-chrono,cadical-trail,cadical-trail-enhanced}
\end{axis}
\end{tikzpicture}

### 2020-02-16 figures/cactus_maple2.tex
fb notes for changes

\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt_data.csv};

\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-chrono-dl,maplelcmdist-trail-dl}
\end{axis}
\end{tikzpicture}

>>>>

\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_trail_noforce_data.csv};
\addplot table [x=index, y=original_time, col sep=comma] {results_maplelcmdist_chrono-dl_ncte-true-opt-1lvl_data.csv};

\legend{maplelcmdist,maplelcmdist-chrono,maplelcmdist-trail,maplelcmdist-trail-enhanced}
\end{axis}
\end{tikzpicture}

### 2020-02-16 figures/trail_diagram.tex
fb notes for changes

\draw (3, 10) rectangle (4, 11);
\draw (3, 9.5) rectangle (4, 10);
\draw (3, 7) rectangle (4, 9.5);
\draw (3, 6.5) rectangle (4, 7);
\draw (3, 5) rectangle (4, 6.5);
\draw (3, 6.5) rectangle (4, 3.5);
\draw (5, 4.5) rectangle (6, 5);
\draw (5, 0) rectangle (6, 4.5);
%\draw (5, 2.5) rectangle (6, 3);
%\draw (5, 0) rectangle (6, 2.5);
\node[align=center] at (3.5, 10.75) {lvl 0};
\node[align=center] at (3.5, 9.75) {$x_{d_1}$};
\node[align=center] at (3.5, 8.5) {$x_1$ \\ $x_2$ \\ . \\ . \\ .};
\node[align=center] at (3.5, 6.75) {$x_{d_2}$};
\node[align=center] at (3.5, 5.25) {$x_j$};
\node[align=center] (B) at (5.5, 4.75) {$l_1$};
\node[align=center] at (5.5, 3.75) {$l_2$ \\ . \\ . \\ .};
%\node[align=center] at (5.5, 2.75) {$l_k$};
\node[align=center] at (5.5, 0.25) {$l_n$};
\node[align=center] at (6.5, 4.75) {top};

\draw[decoration={brace,raise=20pt},decorate] (0.5, 5) -- node[right=21pt] {$L_u$} (0.5, 0);
\draw[decoration={brace,raise=20pt},decorate] (0.5, 11) -- node[right=21pt] {$L_0$} (0.5, 5);
\draw[decoration={brace,raise=20pt},decorate] (3.5, 11) -- node[right=21pt] {$L_0$} (3.5, 5);
\draw[decoration={brace,raise=20pt},decorate] (3.5, 5) -- node[right=21pt] {$L_1$} (3.5, 3.5);
\draw[decoration={brace,mirror,raise=20pt},decorate] (3.5, 11) -- node[left=21pt] {$L_T$} (3.5, 3.5);

>>>>

\draw (3, 10) rectangle (4, 11);
\draw (3, 9.5) rectangle (4, 10);
\draw (3, 7) rectangle (4, 9.5);
\draw (3, 6.5) rectangle (4, 7);
\draw (3, 5) rectangle (4, 6.5);
\draw (3, 6.5) rectangle (4, 3.5);
\draw (5, 4.5) rectangle (6, 5);
\draw (5, 0) rectangle (6, 4.5);
%\draw (5, 2.5) rectangle (6, 3);
%\draw (5, 0) rectangle (6, 2.5);
\node[align=center] at (3.5, 10.75) {lvl 0};
\node[align=center] at (3.5, 9.75) {$x_{d_1}$};
\node[align=center] at (3.5, 8.5) {$x_1$ \\ $x_2$ \\ . \\ . \\ .};
\node[align=center] at (3.5, 6.75) {$x_{d_2}$};
\node[align=center] at (3.5, 5.25) {$x_j$};
\node[align=center] (B) at (5.5, 4.75) {$x_{d_3}$};
\node[align=center] at (5.5, 3.75) {$x_k$ \\ . \\ . \\ .};
%\node[align=center] at (5.5, 2.75) {$l_k$};
\node[align=center] at (5.5, 0.25) {$x_n$};
\node[align=center] at (6.5, 4.75) {top};

\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (0.5, 5) -- node[right=21pt] {$L_u$} (0.5, 0);
\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (0.5, 11) -- node[right=21pt] {$L_0$} (0.5, 5);
\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (3.5, 11) -- node[right=21pt] {$L_0$} (3.5, 5);
\draw[decoration={brace,raise=20pt,amplitude=0.2cm},decorate] (3.5, 5) -- node[right=21pt] {$L_1$} (3.5, 3.5);
\draw[decoration={brace,mirror,raise=20pt,amplitude=0.2cm},decorate] (3.5, 11) -- node[left=21pt] {$L_T$} (3.5, 3.5);

### 2020-02-16 paper.tex
fb notes for changes

\documentclass{article}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[noend]{algcompatible}
\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1,hyphens]{url}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\lstset { %
    language=C++,
    %backgroundcolor=\color{black!5}, % set backgroundcolor
    basicstyle=\footnotesize,% basic font setting
}

\begin{document}

\begin{abstract}
\end{abstract}

\section{Introduction}
The vast majority of modern SAT solvers that are used to solve real-world problems are based on the conflict-driven clause learning (CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after every conflict, where all literals from one or more decision levels become unassigned, before the solver resumes making decisions and performing unit propagations. Traditionally, CDCL solvers would backtrack non-chronologically to the conflict level, which is the second highest decision level remaining in the conflict clause after conflict analysis has resolved away all but one literal from the current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently, however, it has been shown that backtracking chronologically (i.e. backtracking one level only after conflict analysis) \cite{DBLP:conf/lpar/2013, DBLP:conf/sat/NadelR18} can be effective on many instances and was implemented in solvers that won the last two SAT competitions. Although chronological backtracking breaks some of the conventional invariants of CDCL solvers, it has been formalized and proven correct \cite{DBLP:conf/sat/MohleB19}.

@@

\section{Background}
We will assume the reader is familiar with SAT solving and the basics of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use the notion of a trail to represent a sequence of literals corresponding to the current partial assignment. The literals appear on the trail in the order in which they were assigned. Associated with every literal on the trail is a decision level. Decision level 0 represents all literals which are implied by the formula itself and subsequent decisions start the beginning of a new decision level. When a literal is made true by unit propagation, the literal is given its decision level and a reason clause, where the reason clause is the clause that was made unit by propagation in order to imply this literal. Most modern solvers use the two-literal watch scheme to make unit propagation more efficient, where each clause is watched by two non-false literals. When a watcher for a clause becomes false, a replacement is searched for in the clause. If no such replacement exists and the other watcher is true or unassigned, then the clause is unit. If no such replacement exists and the other watcher is also false, then the clause is a conflict and needs to be analyzed to produce a learnt clause.

\section{Trail Saving}
It can be observed that a SAT solver often repeats the same propagations over and over again during execution. Although no two descents down the trail will ever be identical because of clause learning, the solver will often propagate many of the same literals with the same reasons in a similar order to what it had done previously, especially since the order of the decisions usually remains similar. We plan to cache the part of the trail that we are backtracking over and use it to restore some of those literals without propagating them from scratch.

@@

\begin{figure}\center\includegraphics[scale=0.7]{trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}

@@

A potentially quadratic speedup can be realized during the propagation of these literals since they are being propagated as a set, rather than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19, DBLP:journals/jair/Gent13}. Restoring literals from the trail cache automatically not only leads to quicker propagation of future literals, but also means that the propagation routine has less work to do as it searches for new watchers in clauses on watchlists. All of the reasons attached to these literals that are automatically restored will now be satisfied, thus the propagation routine can skip over them rather than having to access each of them potentially multiple times to find they are unit.

@@

One disadvantage that chronological backtracking has compared to trail saving is that it doesn't allow the search as much flexibility to jump out of local neighbourhoods, as the common variable selection heuristics tend to depend on \cite{DBLP:conf/dac/MoskewiczMZZM01, DBLP:conf/sat/2015, DBLP:conf/sat/LiangGPC16}. Because chronological backtracking leaves a large portion of the trail assigned, it will behave similarly to how the solver would behave if it was to force the decisions that do occur within this large portion of literals, whereas trail saving does not force any decisions. As variable selection heuristics become more refined, it is likely that adhering to them will become more crucial to performance.

@@

With more clever data structures the efficiency of restoring the trail could be improved. There are a number of parameters which can be tuned, including how often to refresh the trail, how large the cutoff point for rejecting "long" reasons should be, how long to wait to prune the trail cache if one is appending multiple trail caches together, how far to look ahead on the trail cache for conflicts (in terms of number of decisions or total number of literals). We only experimented with a few different settings of these parameters that made sense, but more careful fine-tuning of these parameters could further enhance the effectiveness of the technique. There are potentially other techniques one could use a trail cache for, including inprocessing steps that involve probing. A look down the trail cache is an incomplete but very fast way to do a probing step, as opposed to propagation. It is also possible that trail saving and chronological backtracking could be combined in some way to produce even better results.
\bibliography{ref}{}
\bibliographystyle{splncs04}
\end{document}

@@

\iffalse
\begin{figure}\includegraphics[scale=0.8]{cactus_cadical.pdf}\caption{\small{Comparison of run times for versions of cadical. cadical is without chronological backtracking, cadical-chrono is with chronological backtracking, cadical-trail is with the first version of trail saving, cadical-trail-2levels-ahead is the trail saving with all enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{cactus_maple2.pdf}\caption{\small{Comparison of run times for versions of MapleSAT. maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-trail-enhanced is with trail saving and all enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{test.pdf}\caption{}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{test2.pdf}\caption{}\end{figure}
\fi
\iffalse
\begin{figure}\includegraphics[scale=0.8]{cactus_maple.pdf}\caption{\small{Comparison of run times for versions of MapleSAT. maplelcmdist is from the 2017 SAT Competition, maplelecmdist-chrono is with chronological backtracking (2018 SAT Competition), maplelcmdist-trail is with the first version of trail saving, maplelcmdist-chrono-dl is with chronological backtracking and duplicate learnts (2019 SAT Competition), maplelcmdist-trail-dl is with trail saving and duplicate learnts}}\end{figure}
\fi

>>>>

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[noend]{algcompatible}
\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\lstset { %
  language=C++,
  % backgroundcolor=\color{black!5}, % set backgroundcolor
  basicstyle=\footnotesize,% basic font setting
}

\renewcommand{\note}[1]{{\color{red}{#1}}}

\title{Trail Saving on Backtrack}
\author{Randy Hickey \and Fahiem Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}

\begin{document}
\maketitle
\begin{abstract}
\end{abstract}

\section{Introduction}
\note{Fix later---also \cite{DBLP:journals/jsat/TakRH11} can't be
  first use of trail savings?}

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that backtracking chronologically
(i.e. backtracking one level only after conflict analysis)
\cite{DBLP:conf/lpar/2013,DBLP:conf/sat/NadelR18} can be effective on
many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

@@

\section{Background}
\note{elaborate backtrack after learning a clause. Also describe
  properties of the trail (a) reason valid every reason clause is made
  unit by its prefix, (b) non-contradictory, (c) non-conflicting (no
  clause is falsified), and (d) propagation complete }

We will assume the reader is familiar with SAT solving and the basics
of the CDCL algorithm \cite{DBLP:series/faia/SilvaLM09}. We will use
the notion of a trail to represent a sequence of literals
corresponding to the current partial assignment. The literals appear
on the trail in the order in which they were assigned. Associated with
every literal on the trail is a decision level. Decision level 0
represents all literals which are implied by the formula itself and
subsequent decisions start the beginning of a new decision level. When
a literal is made true by unit propagation, the literal is given its
decision level and a reason clause, where the reason clause is the
clause that was made unit by propagation in order to imply this
literal. Most modern solvers use the two-literal watch scheme to make
unit propagation more efficient, where each clause is watched by two
non-false literals. When a watcher for a clause becomes false, a
replacement is searched for in the clause. If no such replacement
exists and the other watcher is true or unassigned, then the clause is
unit. If no such replacement exists and the other watcher is also
false, then the clause is a conflict and needs to be analyzed to
produce a learnt clause.

\section{Trail Saving}
\note{
  \begin{enumerate}
  \item Use simple backtrack and save to motivate and state the
      invariant. 
  \item Introduce notation for concatenating two trails.
  \item Name and state the invariant in text (e.g. a
      defn)---concenating the cached trail with the current trail
      yields a trail that is always reason valid, but not necessarily
      non-contradictory, non-conflicting, nor propagation complete).
  \item Proof (simple) that the invariant holds for caching on
      backtrack.
  \item On backtrack at least the asserted literal and any new
      propagants at the backtrack level are added.
  \item Prove that the invariant holds (???do we need an invariant or
      simply a property of the augmented trail) for trails with extras
      inbeween.
      $T_{\mathit{old}} + T_{\mathit{newlyAdded}} +
      T_{\mathit{saved}}$ also satisfies the invariant
  \item re-acheving propagation completeness for the augmented trail
      (a) only add one level of the cached trail at a time, and do
      unit prop after (b) only add non-contradicted decision from the
      old trail. Also note that conflicts are valid and can be learnt
      from even if the augmented trail is not propagation complete. 
  \item Finally figure out a way to formalize the extra technique of
      storing on top after repeated backtracking. 
  \item Explain how freedom for the search is maintained by makeing
      decisions then consulting the trail.
  \item 
  \end{enumerate}
}

It can be observed that a SAT solver often repeats the same
propagations over and over again during execution. Although no two
descents down the trail will ever be identical because of clause
learning, the solver will often propagate many of the same literals
with the same reasons in a similar order to what it had done
previously, especially since the order of the decisions usually
remains similar. We plan to cache the part of the trail that we are
backtracking over and use it to restore some of those literals without
propagating them from scratch.

@@

\begin{figure}\center\includegraphics[scale=0.7]{figures/trail_diagram.pdf}\caption{Storing into a trail cache after backtracking.}\end{figure}

@@

A potentially quadratic speedup can be realized during the propagation
of these literals since they are being propagated as a set, rather
than one at a time, as argued in \cite{DBLP:conf/sat/HickeyB19,DBLP:journals/jair/Gent13}. Restoring literals from the trail cache
automatically not only leads to quicker propagation of future
literals, but also means that the propagation routine has less work to
do as it searches for new watchers in clauses on watchlists. All of
the reasons attached to these literals that are automatically restored
will now be satisfied, thus the propagation routine can skip over them
rather than having to access each of them potentially multiple times
to find they are unit.

@@

One disadvantage that chronological backtracking has compared to trail
saving is that it doesn't allow the search as much flexibility to jump
out of local neighbourhoods, as the common variable selection
heuristics tend to depend on
\cite{DBLP:conf/dac/MoskewiczMZZM01,DBLP:conf/sat/2015,DBLP:conf/sat/LiangGPC16}. Because
chronological backtracking leaves a large portion of the trail
assigned, it will behave similarly to how the solver would behave if
it was to force the decisions that do occur within this large portion
of literals, whereas trail saving does not force any decisions. As
variable selection heuristics become more refined, it is likely that
adhering to them will become more crucial to performance.

@@

With more clever data structures the efficiency of restoring the trail
could be improved. There are a number of parameters which can be
tuned, including how often to refresh the trail, how large the cutoff
point for rejecting "long" reasons should be, how long to wait to
prune the trail cache if one is appending multiple trail caches
together, how far to look ahead on the trail cache for conflicts (in
terms of number of decisions or total number of literals). We only
experimented with a few different settings of these parameters that
made sense, but more careful fine-tuning of these parameters could
further enhance the effectiveness of the technique. There are
potentially other techniques one could use a trail cache for,
including inprocessing steps that involve probing. A look down the
trail cache is an incomplete but very fast way to do a probing step,
as opposed to propagation. It is also possible that trail saving and
chronological backtracking could be combined in some way to produce
even better results.

\bibliography{sat}{}
\bibliographystyle{splncs04}
\end{document}

@@

\iffalse
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_cadical.pdf}\caption{\small{Comparison
        of run times for versions of cadical. cadical is without
        chronological backtracking, cadical-chrono is with
        chronological backtracking, cadical-trail is with the first
        version of trail saving, cadical-trail-2levels-ahead is the
        trail saving with all enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple2.pdf}\caption{\small{Comparison
        of run times for versions of MapleSAT. maplelcmdist is from
        the 2017 SAT Competition, maplelecmdist-chrono is with
        chronological backtracking (2018 SAT Competition),
        maplelcmdist-trail is with the first version of trail saving,
        maplelcmdist-trail-enhanced is with trail saving and all
        enhancements added.}}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/test.pdf}\caption{}\end{figure}
\begin{figure}\includegraphics[scale=0.8]{figures/test2.pdf}\caption{}\end{figure}
\fi
\iffalse
\begin{figure}\includegraphics[scale=0.8]{figures/cactus_maple.pdf}\caption{\small{Comparison
        of run times for versions of MapleSAT. maplelcmdist is from
        the 2017 SAT Competition, maplelecmdist-chrono is with
        chronological backtracking (2018 SAT Competition),
        maplelcmdist-trail is with the first version of trail saving,
        maplelcmdist-chrono-dl is with chronological backtracking and
        duplicate learnts (2019 SAT Competition),
        maplelcmdist-trail-dl is with trail saving and duplicate
        learnts}}\end{figure}
\fi
